---
title: "Practical Machine Learning Course Project"
author: "Timothy M. Rodriguez"
date: "Sunday, June 14, 2015"
output: html_document
---

## Summary
In this project we use the Human Activity Recognition data from [PUC Rio](http://groupware.les.inf.puc-rio.br/har) to build a classifier of the quality of weight lifting movements by a user based on accelerometer data captured from different points on the user's body.

##Exploratory Analysis

First, we load the necessary packages and data.

```{r results='hide'}
require.or.install <- function(package.name) {
  if (!require(package.name, character.only=TRUE)) {
    install.packages(package.name, dep=TRUE)
    if (!require(package.name, character.only=TRUE)) {
      stop(paste("Could not install package ", package.name))
    }
  }
}

download.if.not.present <- function(url, file.name) {
  if (!file.exists(file.name)) {
    download.file(url, file.name)
  }
}

require.or.install("caret")
require.or.install("doParallel")
registerDoParallel(makeCluster(detectCores()))

training.file.name <- "training.csv"
testing.file.name <- "testing.csv"
download.if.not.present("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", "training.csv")
download.if.not.present("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", "testing.csv")

training <- read.csv(training.file.name)
testing <- read.csv(testing.file.name)

```

Next, we do some further analysis to figure out which features are useful.
```{r}
near.zero <- nearZeroVar(training, saveMetrics=TRUE)
print(near.zero[near.zero$nzv==TRUE, ])
```
We see that there are quite a few variables with near zero variance, so we remove them.
```{r}
col.names <- rownames(near.zero[near.zero$nzv==FALSE, ])
sub.training <- subset(training, select=col.names)
sub.testing <- subset(testing, select=col.names[1:99]) #we leave off the classe variable
```
There are also a few diagnostic variables that should not be included for creating a generalizable model such as the user's name, the start or end of windows during data collections, etc.  This is good data to have from a gathering point of view, but is not useful for building a classifier that generalizes to usage outside of a laboratory.
```{r}
sub.training <- subset(sub.training, select=c(-user_name, -raw_timestamp_part_1, -raw_timestamp_part_2, -cvtd_timestamp, -num_window))
sub.testing <- subset(sub.testing, select=c(-user_name, -raw_timestamp_part_1, -raw_timestamp_part_2, -cvtd_timestamp, -num_window))
```
Since our test set is really only used for final grading. We'll split the training data further down into a training and test set.
```{r}
set.seed(7356)
training.ids <- createDataPartition(y=sub.training$classe, p=0.05, list=FALSE)
cleaned.training <- training[training.ids, ]
cleaned.testing <- training[-training.ids, ]
fit <- train(classe ~ ., data=cleaned.training, method="rf")
```
